---
title: "EB Workbook"
author: "Hugo"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(dplyr)
```

# Loading scrapped data 

```{r}
# Prior Data
# See jupyter notebook to see data scraping related code
data <- read.csv("826SR.csv")
print(data)
# Remove first two columns and filter out NA values
cleaned_data <- data[, c(3:5)] %>% 
  filter(!is.na(`SuccessRate`))

names(cleaned_data) <- c('PK', 'PKatt','SR')

#Filter out players with less than 5 attempts 
cleaned_data <- cleaned_data %>%
  filter(`PKatt` >= 5)
#Convert 1.0 to 0.9999 to be able to fit a beta distribution
cleaned_data <- cleaned_data %>%
  mutate(SR = ifelse(SR == 1.0, 0.9999, SR)) 

cleaned_data <- cleaned_data %>%
  mutate(SR = ifelse(SR == 0, 0.0001, SR)) 

cleaned_data
  
```



```{r}
hist(cleaned_data$SR, main = "conversion rate", breaks = 10 )
```

```{r}
mu_hat = mean(cleaned_data$SR)
print(paste0("mu_hat = ", mu_hat))
n_i = length(cleaned_data$SR)
print(paste0("n_i = ", n_i))
sigma_hat2 = var(cleaned_data$SR) - (mu_hat*(1-mu_hat))/n_i
print(paste0("sigma_hat = ", sigma_hat2))

a0 = mu_hat*((mu_hat*(1-mu_hat))/sigma_hat2 -1)
print(paste0("a0 = ", a0))
b0 = (1-mu_hat)*((mu_hat*(1-mu_hat))/sigma_hat2 -1)
print(paste0("b0 = ", b0))

```
```{r}
plot(function(x) dbeta(x, shape1 = a0, shape2 = b0), 
    from = 0, to = 1, 
    main = "Estimated Beta Distribution (prior success rate)",
    ylab = "Density",
    xlab = "x")

```
```{r}
hist(cleaned_data$SR, freq = FALSE, main = "Empirical Bayes prior Beta Distribution",
     xlab = "Value")
lines(x <- seq(0, 1, length = 100), 
      dbeta(x, shape1 = a0, shape2 = b0), 
      col = "red", lwd = 2)
legend("topleft", legend = paste("Beta(", round(a0,2), ", ", round(b0,2), ") distribution", sep = ""), 
       col = "red", lwd = 2)
```
The fit of the estimated beta distribution is ok, considering the relatively non regular histogram. The spikes in the histogram can be explained by a bias in the data towards specific success rate values, such as when the number of trials in small (ex: 3/5 = 0.6, 4/5 = 0.8, 5/5, = 1, no in between values). Here the choice of 5 attempts minimum was made follow the choices made in the football website article, but note that the number of data points decreases logarithmically with the minimum number of attempts (see plot in jupyter notebook).

```{r}

#library(MASS)
#plot(density(cleaned_data$SR), main = "Oberved Data density Distribution")

# Alternative to estimate a0 and b0
#m <- fitdistr(cleaned_data$SR, "beta",
              #start = list(shape1 = a0, shape2 = b0))

#alpha0 <- m$estimate[1]
#beta0 <- m$estimate[2]
#print(paste0("alpha0 = ", alpha0))
#print(paste0("beta0 = ", beta0))
```


```{r}
# Posterior calculations

# Example 1: Hypothetical situation with worst than average players.

library(ggplot2)
library(coda) # For HPDI calculation

# EB prior parameters
prior_alpha <- a0
prior_beta <- b0

# Hypothetical player data, both having 0.6 Conversion Rate
# Recall league average mu_hat = 0.813
player_data <- list(
  A = list(successes = 6, trials = 10),
  B = list(successes = 30, trials = 50)
)

# Calculate posterior parameters, credible intervals, and HPDI
results <- list()
meanvar <- list()
for (player in names(player_data)) {
  successes <- player_data[[player]]$successes
  trials <- player_data[[player]]$trials
  fails <- trials - successes
  # Posterior parameters (From Equation (5) result)
  posterior_alpha <- prior_alpha + successes
  posterior_beta <- prior_beta + fails
  
  # posterior mean and variance
  posterior_mean <- (prior_alpha + successes) / (prior_alpha + prior_beta + trials) #Equation (6)
  posterior_variance <- (prior_alpha + successes)*(prior_beta + fails) / ((prior_alpha + prior_beta + trials)^2 * (prior_alpha + prior_beta + trials + 1)) # Equation (7)
  
  # 95% credible interval
  credible_interval <- qbeta(c(0.025, 0.975), posterior_alpha, posterior_beta)
  
  # Simulate posterior
  simulated_posterior <- rbeta(10000, posterior_alpha, posterior_beta)
  
  # 95% HPDI
  hpdi <- HPDinterval(as.mcmc(simulated_posterior), prob = 0.95)
  
  results[[player]] <- list(
    posterior_alpha = posterior_alpha,
    posterior_beta = posterior_beta,
    credible_interval = credible_interval,
    hpdi = hpdi,
    simulated_posterior = simulated_posterior
  )
  meanvar[[player]] <- list(posterior_mean = posterior_mean, posterior_variance = posterior_variance)
  
}
print(meanvar)
#Plotting:
x <- seq(0, 1, length.out = 1000)
density_data <- data.frame(
  x = rep(x, times = 2),
  density = c(
    dbeta(x, results$A$posterior_alpha, results$A$posterior_beta),
    dbeta(x, results$B$posterior_alpha, results$B$posterior_beta)
  ),
  player = rep(c("A", "B"), each = length(x))
)

interval_data <- data.frame(
  player = c("A", "B"),
  lower_credible = sapply(results, function(r) r$credible_interval[1]),
  upper_credible = sapply(results, function(r) r$credible_interval[2]),
  lower_hpdi = sapply(results, function(r) r$hpdi[1]),
  upper_hpdi = sapply(results, function(r) r$hpdi[2])
)
print(interval_data)


# Plot-generating code written with the help of chatgpt.com

# Plot the posterior distributions with intervals
ggplot(density_data, aes(x = x, y = density, color = player)) +
  geom_line(size = 1) +
  geom_vline(data = interval_data, aes(xintercept = lower_credible, color = player), linetype = "dashed") +
  geom_vline(data = interval_data, aes(xintercept = upper_credible, color = player), linetype = "dashed") +
  geom_vline(data = interval_data, aes(xintercept = lower_hpdi, color = player), linetype = "dotted") +
  geom_vline(data = interval_data, aes(xintercept = upper_hpdi, color = player), linetype = "dotted") +
  geom_vline(xintercept = mu_hat, color = "black", linetype = "solid", size = 1, alpha = 0.7) + # League average
  geom_vline(xintercept = 0.6, color = "purple", linetype = "solid", size = 1, alpha = 0.7) + # Players SR
  labs(
    title = "Posterior Distributions with Credible Intervals (dashed) and HPDI (dotted)",
    x = "Success Rate",
    y = "Density"
  ) +
  scale_color_manual(values = c("A" = "blue", "B" = "red")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(title = "Player"))


```

While both players have below average Conversion rate to begin with (purple vetical line), it can be observed that player A with the least amount of attempted penalty kick has his posterior density more easily shifted to the league average (Black vertical line). However, both the credible interval (dashed lines) and the HPDI (dotted lines) are much wider for player A and includes a SR region to the left of Player B, which means that Player B could still have a SR worst than player B. Player B's intervals on the other hand are much tighter, indicating more certainty in the player possible SR.

```{r}
# Example 2: "Perfect" beginner vs. "Above average" experimented.

# EB prior parameters
prior_alpha <- a0
prior_beta <- b0

# Hypothetical player data, 
# Player A with perfect success rate but low amount of attempts
# Player B with slightly lower success rate but higher amount of attempts
# Recall league average mu_hat = 0.813 SR
player_data <- list(
  A = list(successes = 2, trials = 2), # 1.0 SR
  B = list(successes = 85, trials = 100) # 0.85 SR
)

results <- list()
meanvar <- list()
for (player in names(player_data)) {
  successes <- player_data[[player]]$successes
  trials <- player_data[[player]]$trials
  fails <- trials - successes
  # Posterior parameters
  posterior_alpha <- prior_alpha + successes
  posterior_beta <- prior_beta + fails
  
   posterior_mean <- (prior_alpha + successes) / (prior_alpha + prior_beta + trials) #Equation (6)
  posterior_variance <- (prior_alpha + successes)*(prior_beta + fails) / ((prior_alpha + prior_beta + trials)^2 * (prior_alpha + prior_beta + trials+1)) # Equation (7)
  
  # 95% credible interval
  credible_interval <- qbeta(c(0.025, 0.975), posterior_alpha, posterior_beta)
  
  # Simulate posterior
  simulated_posterior <- rbeta(10000, posterior_alpha, posterior_beta)
  
  # 95% HPDI
  hpdi <- HPDinterval(as.mcmc(simulated_posterior), prob = 0.95)
  
  results[[player]] <- list(
    posterior_alpha = posterior_alpha,
    posterior_beta = posterior_beta,
    credible_interval = credible_interval,
    hpdi = hpdi,
    simulated_posterior = simulated_posterior
  )
  meanvar[[player]] <-list(posterior_mean = posterior_mean, posterior_variance = posterior_variance)
}

print(meanvar)
#for plotting:
x <- seq(0, 1, length.out = 1000)
density_data <- data.frame(
  x = rep(x, times = 2),
  density = c(
    dbeta(x, results$A$posterior_alpha, results$A$posterior_beta),
    dbeta(x, results$B$posterior_alpha, results$B$posterior_beta)
  ),
  player = rep(c("A", "B"), each = length(x))
)

interval_data <- data.frame(
  player = c("A", "B"),
  lower_credible = sapply(results, function(r) r$credible_interval[1]),
  upper_credible = sapply(results, function(r) r$credible_interval[2]),
  lower_hpdi = sapply(results, function(r) r$hpdi[1]),
  upper_hpdi = sapply(results, function(r) r$hpdi[2])
)
print(interval_data)

# Plot-generating code written with the help of chatgpt.com
# Plot the posterior distributions with intervals
ggplot(density_data, aes(x = x, y = density, color = player)) +
  geom_line(size = 1) +
  geom_vline(data = interval_data, aes(xintercept = lower_credible, color = player), linetype = "dashed") +
  geom_vline(data = interval_data, aes(xintercept = upper_credible, color = player), linetype = "dashed") +
  geom_vline(data = interval_data, aes(xintercept = lower_hpdi, color = player), linetype = "dotted") +
  geom_vline(data = interval_data, aes(xintercept = upper_hpdi, color = player), linetype = "dotted") +
  geom_vline(xintercept = mu_hat, color = "black", linetype = "solid", size = 1, alpha = 0.7) + # League average
  #geom_vline(xintercept = 0.6, color = "purple", linetype = "solid", size = 1, alpha = 0.7) + # Players SR
  labs(
    title = "Posterior Distributions with Credible Intervals (dashed) and HPDI (dotted)",
    x = "Success Rate",
    y = "Density"
  ) +
  scale_color_manual(values = c("A" = "blue", "B" = "red")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(title = "Player"))

```

